Evolution of ORM Systems: Formal Foundations

### 2013-01-03 doc/description.tex
added manipulators for database structure

@@

\begin{center}
$\rho_{c}: Class \times Database \rightarrow Database $ \\ 
$\rho_{c}(c,d) = tables(d) \cup table(name(c), primaryKey("Id"), \varnothing, \varnothing) $
\end{center}

@@

\subsection{Software Creation}
There are defined three transformations which are able to create a software: $newClass$, $addProperty$, $addAssociation$. These three transformations can be used to create a software from a scratch. The first of them has no impact on stored data, whereas the rest of them can in some cases.
\begin{gather*}
\mathbf{addProperty} : Label \times Property \times State \rightarrow State \\
\Psi(s, addProperty(l, p)) = application(s) \circ addProp(l,p) \\
\Phi(s, addProperty(l, p)) = database(s) \bullet insertProp(l, p) \\
s \in Softwares, l \in Label, p \in Properties
\end{gather*}
Adding a property into an application structure has to verify a name collision only:
\begin{gather*}
addProp : Label \times Property \times Application \rightarrow Application \\
addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p } \\
l \in Label, p \in Properties, a \in Applications
\end{gather*}
Whereas the $insertProp$ on database level can change stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{gather*}
insertProp(p, d) = \begin{cases}
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge (!\;mandatory(p) \vee !\;instantiated(owningClass(p))}{ columns(t) \cup \rho(p,d)} 
\\ \\ 
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered} columns(t) \cup 
 \rho(p,d) \\ \wedge \forall r \in data(d), reference(r) = name(t) : pairs(r) \\ \cup [name(p), defaultValue(p)]
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
tables(d) \cup  \rho(p,d) \wedge \forall r \in row(owningClass(p)) : data(d)\\ \cup row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))]) 
\end{gathered}
} 
\end{cases}\\
p \in Properties, d \in Dataabses
\end{gather*}

@@

The $insertAssoc$ function is defined as:
\begin{gather*}
insertAssoc : Association \times Database \times (Row \times Row) \\
insertAssoc(a, d, f_m) = \begin{cases}
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
columns(t_s) \cup \rho(a,d) \\ \wedge \forall r \in f_m(owningClass(a), reference(a)) : \\ data(t_t) \cup row(name(t_s), [name(a), \\ value(primaryKey(first(r)))]) 
\end{gathered}
 }
 \\ \\
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) > 1}{\begin{gathered} tables(d) \cup
\rho(a,d) \\ \wedge \forall r \in f_m(owningClass(a), reference(a)),\\ t \in tables(d) : name(t) = name(a) : \\ data(t) \cup row(name(a), [name(t_s), \\ value(primaryKey(first(r)))] [name(t_t), \\ value(primaryKey(second(r)))])
\end{gathered}
 } 
 \end{cases}\\
 a \in Associations, d \in Databases
\end{gather*}

\begin{gather*}
\mathbf{newClass} : Class \times State \rightarrow State \\
\Psi(s, newClass(c)) = application(s) \circ addClass(c) \\
\Phi(s, newClass(c)) = database(s) \bullet insertClass(c) 
\end{gather*}
\begin{gather*}
addClass: Class \times Application \rightarrow Application \\ 
addClass(c, a) = \inference{\forall s \in classes(a) : name(s) \neq name(c)}{classes(a) \cup c} \\ \\
c \in Classes, a \in Applications
\end{gather*}
\paragraph{Add Class} The $insertClass$ operation changes the structure of the database schema only, because the $insertAssoc$ transformation needs a mapping function to work properly, which is not part of the $newClass$ definition. The situation reflect the modeling praxis when the structure of a class is defined and then it is associations are created.
\begin{gather*}
insertClass : Class \times Database \rightarrow Database \\
insertClass(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	%\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) \\ 
	assocaiations(c) = \varnothing
}{
\begin{gathered}
  tables(d) \cup \rho(c) \\ \wedge \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ 
\end{gathered}
}
\end{gather*}

@@

\subsubsection{Altering a Software}	
\paragraph{Copy Property}
All transformations described earlier does change only the structure of a application. These operation can be used in case there are no instances in the application. If instances are present in the application, the decomposition transformations could cause a data lost. Therefore we define  the transformation for copying a property from one class to another. The copy property is the basic way to copy data from instance of a class to instance of another class.

The copy property can be implemented as a transformation composed by operations for adding and removing property, nevertheless this composition would not work with instances, thus we define the copy property operation as a basic and atomic operation which will help us to handle instances.

>>>>

\subsection{Transformations for Manipulating a Database}
The database is composed from two separated parts - database schema and data. Several transformation are defined to make manipulating a database easily in this paper.

\begin{gather*}
addColumn : Label \times Column \times Database \rightarrow Database \\ addColum(l, c, d) = \begin{cases}
columns(t) \cup c \; \mathbf{if} t \in tables(d) \wedge name(t) = l \\ \;\;\; \wedge \forall f \in columns(t) : name(f) \neq name(c)
\\ 
Inconsistent \;d 
 \end{cases}
\\ l \in Label, c \in Columns, d \in Databases
\end{gather*}

\begin{gather*}
addTable : Table \times Database \rightarrow Database \\ 
addTable(t, d) = \begin{cases}
tables(d) \cup c \; \mathbf{if} \forall e \in tables(d) \wedge name(t) \neq name(e)
\\ 
Inconsistent \;d 
 \end{cases}
\\ t \in Tables,  d \in Databases
\end{gather*}

@@

\begin{center}
$\rho_{c}: Class \times Database \rightarrow Database $ \\ 
$\rho_{c}(c,d) = table(name(c), primaryKey("Id"), \varnothing, \varnothing) $
\end{center}

@@

\subsection{Software Creation}
There are defined three transformations which are able to create a software: $newClass$, $addProperty$, $addAssociation$. These three transformations can be used to create a software from a scratch. The first of them has no impact on stored data, whereas the rest of them can in some cases.
\begin{gather*}
\mathbf{addProperty} : Label \times Property \times State \rightarrow State \\
\Psi(s, addProperty(l, p)) = application(s) \circ addProp(l,p) \\
\Phi(s, addProperty(l, p)) = database(s) \bullet insertProp(l, p) \\
s \in Softwares, l \in Label, p \in Properties
\end{gather*}
Adding a property into an application structure has to verify a name collision only:
\begin{gather*}
addProp : Label \times Property \times Application \rightarrow Application \\
addProp(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in  properties(p) : name(q) \neq name(p)}{properties(c) \cup p } \\
l \in Label, p \in Properties, a \in Applications
\end{gather*}
Whereas the $insertProp$ on database level can change stored data too, because a mandatory property can be added into a class with already existing instances, then a default value has to be added to all existing instances.
\begin{gather*}
insertProp(p, d) = \begin{cases}
\inference{!\:mandatory(p) \vee !\:instantiated(owningClass(p))}{addColumn(l, \rho(p,d), d)} 
\\ \\ 
\inference{ \wedge mandatory(p) \vee instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered} addColumn(l, \rho(p,d), d) \\ \bullet \forall r \in data(d), reference(r) = name(t) : pairs(r) \\ \cup [name(p), defaultValue(p)]
\end{gathered}
} 
\\ \\
\inference{ \exists t \in tables(d) : name(t) = name(owningClass(p)) \\ \wedge mandatory(p) \wedge instantiated(owningClass(p)) \\ \wedge cardinality(p) \leq 1}{\begin{gathered}
addTable(\rho(p,d), d) \bullet \forall r \in row(owningClass(p)) : data(d)\\ \cup row(name(p), ["value", defaultValue(p)], \\ [name(p), value(primaryKey(r))]) 
\end{gathered}
} 
\end{cases}\\
p \in Properties, d \in Dataabses
\end{gather*}

@@

The $insertAssoc$ function is defined as:
\begin{gather*}
insertAssoc : Association \times Database \times (Row \times Row) \\
insertAssoc(a, d, f_m) = \begin{cases}
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) \leq 1}{\begin{gathered}
addColumn(name(t_s), \rho(a,d), d) \\ \bullet \forall r \in f_m(owningClass(a), reference(a)) : \\ data(t_t) \cup row(name(t_s), [name(a), \\ value(primaryKey(first(r)))]) 
\end{gathered}
 }
 \\ \\
\inference{ \exists t_s, t_t \in tables(d) : name(t_s) = name(owningClass(a)) \\ \wedge name(t_t) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_t) : name(f) = name(a) \\ startCardinality(a) > 1}{\begin{gathered} addTable(\rho(a,d),d) \\ \bullet \forall r \in f_m(owningClass(a), reference(a)),\\ t \in tables(d) : name(t) = name(a) : \\ data(t) \cup row(name(a), [name(t_s), \\ value(primaryKey(first(r)))] [name(t_t), \\ value(primaryKey(second(r)))])
\end{gathered}
 } 
 \end{cases}\\
 a \in Associations, d \in Databases
\end{gather*}

\begin{gather*}
\mathbf{newClass} : Class \times State \rightarrow State \\
\Psi(s, newClass(c)) = application(s) \circ addClass(c) \\
\Phi(s, newClass(c)) = database(s) \bullet insertClass(c) 
\end{gather*}
\begin{gather*}
addClass: Class \times Application \rightarrow Application \\ 
addClass(c, a) = \inference{\forall s \in classes(a) : name(s) \neq name(c)}{classes(a) \cup c} \\ \\
c \in Classes, a \in Applications
\end{gather*}
\paragraph{Add Class} The $insertClass$ operation changes the structure of the database schema only, because the $insertAssoc$ transformation needs a mapping function to work properly, which is not part of the $newClass$ definition. The situation reflect the modeling praxis when the structure of a class is defined and then it is associations are created.
\begin{gather*}
insertClass : Class \times Database \rightarrow Database \\
insertClass(c, d) = \inference{ \forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	%\forall a \in associations(c), \exists u \in  tables(d) : reference(a) = name(u) \\ 
	assocaiations(c) = \varnothing
}{
\begin{gathered}
 addTable(\rho(c, d),d) \\ \wedge t \rho(singleProperties(c)) \wedge \rho(collectionProperties(c)) \\ 
\end{gathered}
}
\end{gather*}

@@

\paragraph{Copy Property}
All transformations described earlier does change mostly the structure of a application only. Therefore we define  the transformation for copying a property from one class to another. The copy property is the basic way to copy data from instance of a class to instance of another class.

The copy property can be implemented as a transformation composed by operations for adding and removing property, nevertheless this composition would not work with instances, where the $copyData$ transformation is used.
